---
title: "Statistical Inference Project Part 1: Simulation Exercise"
author: "Jared Kramer"
date: "1/22/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

In this report, I investigate the statistical properties of the mean
of 40 random variables drawn from an exponential distribution.   Using 1,000 simulations, 
I investigate the *mean* of the distribution of these means (comparing them to the 
theoretical mean), as well as the *variance* of this distribution (comparing the sample 
variance to the theoretical variance).   Finally, I observe the shape of the distribution
compared to a normal distribution. 

### Simulations

First, I set parameters to be used in the simulations: lambda will be 0.2 for 
all simulations.   The parameter `Nexp` is set to 40 to indicate the number of exponential
random variables that will be generated in each simulation, and `Nsimul` indicates that
there will be 1,000 simulations run. 

```{r params}
set.seed(1234)
lambda <- 0.2; Nexp <- 40; Nsimul <- 1000
```

For the simulations, I draw `r Nsimul` * `r Nexp` random numbers from an exponential 
distribution, and reshape into a matrix with 1000 rows and 40 columns.
Each row is thus a single simulation of 40 data points. I then use `apply` to compute 
the mean of each row/simulation. 

```{r simul}
M <- matrix(rexp(Nsimul * Nexp, rate = lambda), nrow = Nsimul, ncol = Nexp)
means <- apply(M, 1, mean)
```

### Sample vs. Theoretical Means

Theoretically, the mean of 40 random numbers drawn from a single distribution 
should have a mean that equals the mean of the  underlying distribution.
Since the underlying distribution (here, exponential) has mean `1/lambda` 
(here,`r 1/lambda`)  the distribution of the means should also have mean `1/lambda`, 
i.e., `r 1/lambda`.

In our simulations, the sample mean is very close to this theoretical value, i.e.,: 
`r mean(means)` vs. the theoretical of `r 1/lambda`. We can see graphically from the
below histogram that this is the case as well (red vertical line shows sample mean, 
and blue vertical line shows theoretical).

```{r means, fig.height = 4}
hist(means, main = "Means of Simulations of 40 Exponential Data Points", 
     xlab = "Simulation Means", breaks = 20)
abline(v = mean(means), col = "red", lwd = 2)
abline(v = 1/lambda, col = "blue", lwd = 2)
legend("topright", legend = c("Sample Mean", "Theoretical Mean"),
       col = c("red", "blue"), lty = 1)
```

### Sample vs. Theoretical Variance

Theoretically, the distribution of the mean of 40 random numbers drawn from a single 
distribution should have a variance that equals the variance of the underlying 
distribution, *divided by* the number of data points contributing to the mean. 

In this case, the underlying distribution is exponential with standard deviation
`1 / lambda` (i.e., `r 1/lambda`) and variance `1 / lambda ^ 2` (i.e., `r 1/lambda^2`), 
the theoretical  variance of the means of each simulation is `1 / (lambda ^ 2 * Nexp)`, 
or `r 1 / (lambda ^ 2 * Nexp)`.  The simulated variance is quite close to this value,
at, `r var(means)`.

We can visualize this variability differential by again plotting a histogram of the 
simulation means, and overlaying a normally distributed curve with mean equal to the 
simulated mean, and variance equal to the theoretical variance.   

```{r variance graph, fig.height = 3}
hist(means, main = "Comparison of Variability of Means", prob = TRUE, breaks = 20, 
     xlab = "Simulation Means")
x <- seq(min(means), max(means), length = 100)
y <- dnorm(x, mean = mean(means), sd = 1/(lambda * sqrt(Nexp)))
lines(x, y)
```

### Shape of Distribution 

For a large number of data points per simulation, the central limit theorem would imply 
that the distribution of the means will approach the shape of a normal distribution.  

We can visualize this roughly from the histogram in the previous section, which seems
to match reasonably well with the overlaid normal density curve.  For a more precise 
comparison, I use a quantile-quantile plot that compares the quantiles of a normalized
version of the means (subtracting the sample mean from each and dividing by the 
sample standard deviation), with the theoretical quantiles of a standard normal 
distribution. 

The graph shows a reasonably good match throughout the bulk of the distribution, although
in the "tails" of the the distribution the simulations diverge somewhat from normal: 
on the left tail, the simulated distribution is slimmer than in a normal distribution, 
and in the right tail, the simulated distribution is slightly "fatter".   

```{r shape, fig.height = 3}
normalized <- (means - mean(means))/(sd(means))
qqnorm(normalized); qqline(normalized)
```

